A comprehensive analysis of natural sequence variation within the <i>Arabidopsis thaliana</i> nuclear auxin signaling pathway

### 2018-03-07 manuscript/manuscript.Rmd
Romi's AFB section

# Results  
## TIR1/AFB genes  
Romi
[@gray_identification_1999; @kepinski_auxininduced_2004; @tan_mechanism_2007; @dezfulian_oligomerization_2016; @wu_evolution_2017b]

## Aux/IAA genes

@@

## TPL/TPR genes  
  The Auxin signaling pathway utilizes the TOPLESS (TPL) family of Gro/TLE/TUP1 type co-repressor proteins to maintain auxin responsive genes in a transcriptionally-repressed state in the absence of auxin (cite). In Arabidopsis the TPL family includes TOPLESS and four TPL-related proteins (TPR1-4), which are comprised of three structural domains: an N-terminal TPL domain and two beta-propellar domains. TPL is recruited to the AUX/IAA proteins through interaction with a conserved protein motif known as an Ethylene-responsive element binding factor-associated amphiphilic repression (EAR) domain (cite). Canonical EAR domains have the amino acid sequence LxLxL, as is found in domain I of the AUX/IAAs (figure xA, cite). TPL co-repressors bind EAR domains through direct interaction with the TPL N-terminal C-terminal to LisH (CTLH) domain (citations of pre-structure founding papers/reviews). Recent structural analyses of the TPL N-terminal domain have highlighted the precise interaction interface between TPL and AUX/IAA EAR domains, as well as the TPL-TPL dimerization and tetramerization motifs (Avarelilo et al 2017, Ke et al, 2015), and even the residues required for higher-order multimers of TPL tetramers (Ma et al, 2017). The two tandem beta propellar domains of TPL/TPRs have as yet not been crystallized, however numerous high resolution crystal structures of beta propellar domains allow for structural prediction of the TPL c-terminal beta propellar domains and the locations of likely interaction faces (Figure, X).

  The TOPLESS co-repressor family generally exhibits a high level of sequence conservation at the amino acid sequence level across the Arabidopsis accessions with all ${\pi_N/\pi_S}$ values below 1 (Table X). The closely related TPL and TPR1 have the highest ${\pi_N/\pi_S}$ values (TPL-0.425, TPR1-0.739), suggesting that these these two related genes tolerate a higher degree of sequence diversity compared to TPR2-4. The TPL/TPR family demonstrates a high level of sequence conservation across the N-terminal TPL domain, with the lowest of observed ${\pi_N/\pi_S}$ values (Figure X). Indeed, all non-synonymous mutations observed in this region are either in linker domains between alpha helices, or are highly conservative mutations within helices (i.e. Valine to Isoleucine), which would be predicted to exhibit little effect on TPL folding and activity.

  The increased N-terminal conservation underscores the importance of the N-terminal TPL domain. For example, the initial tpl-1 mutation (N176H) in the ninth helix is a dominant gain-of-function allele (Long et al, 2002, PMID12050130), which is capable of binding wild-type TPL protein and inducing protein agreggation (Ma, 2017, pmid28630893). It is therefore understandable that very little diversity is observed in the ninth alpha helix of TPL/TPRs as mutations in this domain could act in a dominant negative fashion, rendering any de novo mutation in this region embryo lethal. However, the entire N-terminal domain exhibits very low rates of nucleotide diversity, suggesting that each domain in this region is critical at the population level. Indeed, the crystal structure of TPL has revealed that the TPL domain is responsible for the multiple roles of TPL/TPR proteins: The LisH domain and CRA domain coordinate TPL dimerization and tetramerization respectively (26601214, 28698367), the CTLH domain is critical for interaction with DNA-binding transcription factors (26601214), and helix 9 within the CRA domain may be critical for the formation of higher-order TPL complexes (multimers of TPL tetramers, 28630893).

  We then compared the frequency of SNPs that cause nonsynonymous mutations across the TPL domains. Compared to the TPL domain, both the proline rich and WD-40 domains exhibit a higher rate of missense mutations accross both of the WD-40 repeat regions, here labeled WD40-1 and WD40-2 respectively. We utilized the Phyre2.0 modeling software to generate a rough model of the TPL/TPR WD-40 repeats stricly for the purpose of mapping non-synonymous mutations onto a visible protein chassis (Figure X). Here we can see both faces of each predicted beta-propellar domains, which are connected by a short linker region. Through utilization of protein modelling, it becomes clear that we observe conservation of amino acid sequence at the center of the beta-propellar domains. These observations could be used as the basis for forward mutagensis strategies to test whether this region of TPL functions similarly to its homolog TLE, which binds its target genes EH1 and the C-terminal WRPW/Y motif (Hairy/Hes/Runx) in the center of the beta-propeller domain (pmid 16762837).

  It is interesting to note that despite the presence of 5 TPL/TPR genes in the genome, which are considered to be redundant (pmid 16763149), very few SNPs are observed that are predicted to result in loss of function for any one TPL/TPR gene. In fact, only one deletion was observed (665delG) that creates a frameshift mutation in TPR4 in a single accession (Aiell-4 in Italy). This mutation occurs in codon 222, just downstream of the TPL N-terminal domain, and would generate an early stop codon 11 codons downstream of the deletion site. It is unclear whether this N-terminal fragment would be expressed at any appreciable level, as it is possible that the generation of an early stop codon would trigger nonsense-mediated decay. However, if expressed, it would be fascinating to examine whether this naturally occuring N-terminal truncation retains activity as a co-repressor. Given the observation that the TPL N-terminal domain is sufficient to bind histones (28630893), and repress transcription via synthetic assays in yeast (24979769), it is possible that this represents a semi-functional TPL variant. There has been no evidence for priotization of individual TPL/TPR family members in the auxin response pathway, but experiments modulating the gene copy number of TPR1 revealed the importance of maintaing the proper dosage of TPL/TPR in order to accurately trigger the salicyclic acid innate immune response (20647385). These results, taken together with the absence of TPL/TPR loss-of-function alleles in the Arabidopsis accessions suggest that perhaps the entire family is being conserved at the nucleotide level in the population to meet a gene dosage requirement, with little opportunity for neo- or sub-functionalization.

>>>>

# Results  
## TIR1/AFB genes  
Romi
[@gray_identification_1999; @kepinski_auxininduced_2004; @tan_mechanism_2007; @dezfulian_oligomerization_2016; @wu_evolution_2017b]
@Terrile_Nitric_2012 nitrosylation residues C140 important in SCFTIR1 dimerization and interaction with Aux/IAAs
@Wright_Insights_2017 residues in TIR1 and AFB2 relevant in their degradation of Aux/IAAs.
@Yu_Untethering_2015 E12K E15K residues important in TIR1s interaction with CUL1 (SCF complex).
@tan_mechanism_2007 R401 R484 K485 K74 K113 R114 R403 R344 All important residues binding phytate.
R403 hydrogen bonds with auxin compounds
E487 R489 I488 residues grab the non-polar rings of auxin in place. convex wall part.
I107 P409 F465 F380 C405 P347 P350 F351 mostly non-polar residues interacting with IAAs. Part of Loop 14, 12, and 2.
G142A, F143A, I151A, V117A and G147A mutants dimerization

Auxin acts by binding to receptors that in turn target co-repressors for degradation. Auxin receptors (six in the model plant Arabidopsis thaliana) evolved through gene duplication and diversification early in the history of vascular plants @Parry_Complex_2009. The Nemhauser lab has recently shown that the rate of co-repressor degradation is determined by the identity of both the receptor and co-repressor @Havens_Synthetic_2012, and that this rate sets the pace of development3. 
Although it is unknown whether different receptors play different roles at the organismal level, a number of studies have pointed out differences in the ways the six different receptors in A. thaliana differ in their expression. The expression of AFB5 is induced in the presence of auxin, making this a circadian-regulated gene @Covington_Circadian_2007. It also appears AFB3 is more highly expressed in the roots in the presence of nitrate, allowing increased lateral root formation @Vidal_Nitrate_2010. Although TIR1/AFBs are expressed ubiquitously in A. thaliana tissues, TIR1, AFB2, and AFB3 have been shown to accumulate in the shoot and root meristems and leaf tissues, with slightly different expression patterns in TIR1 @Parry_Complex_2009. AFB1 has been found to have a mutation drastically reducing its ability to assemble into an SCF complex, contributing to it’s increased accumulation, as well as preventing it from degrading Aux/IAAs. Higher order TIR1/AFB mutants show AFB1 to have inconclusive organismal function. However, AFB1 is shown to be able to accumulate ubiquitously in all tissues @Yu_Untethering_2015. When it comes to binding auxins, AFB4 and AFB5 have been shown to preferentially and functionally bind the synthetic auxin picloram, while others have a preference for naturally occurring auxin 3-IAA @Prigge_Arabidopsis_2016. Interestingly, the ways in which TIR1/AFBs are able to bind and mark Aux/IAAs for degradation are variable. TIR1 and AFB2 have slightly different speed at which they perform this function, with AFB2 being generally quicker, as shown by our lab’s yeast degradation system @Havens_Synthetic_2012. Finally, higher order TIR1/AFB mutants show AFB1 to have inconclusive organismal function.
  Using the Natural Variation Webtool developed in our lab, I will characterize the natural variation in the TIR1/AFB family of auxin receptors by looking at the genomes of 1001 accessions of A. thaliana. This webtool allows me to	quantify and compare the nucleotide diversity of different genes in this family, as well as find relevant potentially function-altering variants. 
Using its diversity statistic function has shown me TIR1 and AFB1 both seem to be low diversity, hinting at their likely functional importance. AFB3 and 4 appear to be higher diversity, while their sisters are more conserved. This matches our current understanding of AFB3 and 4 as being minor players in the auxin signaling pathway, but brings up questions about the function of AFB1, which is inconclusive. Using this webtool, the AFB1 sequence can be aligned to the highly characterized TIR1 AA sequence showing locations of known non-synonymous variants. Although the most important functional regions do not contain variants in AFB1, there is a low diversity mutant, only found in the Can-1 accession, which is in the highly conserved TIR1/AFB oligomerization domain. Because variants eliciting significant changes in function often are not selected for, the fact that this variant is extremely low diversity holds promise that it is associated with a strong auxin signaling phenotype. I will order and characterize this accession in order to identify potential auxin signaling phenotypes. This would support that AFB1 functions through oligomerization with other TIR1/AFBs.
  The AFB4 and AFB4 receptors are known to have an extension in their N-terminal domains. Using the alignment feature on this webtool has indicated this region to be the highest in diversity in these proteins, containing lots of missense variants (Fig. 4). This suggests that this extension does not play an important functional role in these proteins.

Figure 4: Alignment of A. thaliana TIR1/AFB F-box domains showing variants. Cyan marks the Fbox binding domain. Yellow, green, and magenta, mark missense and synonymous, synonymous only, and missense only variant locations respectively. Yellow squared E’s show Cul1-interacting residues, differing in AFB1.

@Parry_Complex_2009
1.	Parry G, Calderon-Villalobos LI, Prigge M, et al. Complex regulation of the TIR1/AFB family of auxin receptors. Proceedings of the National Academy of Sciences of the United States of America. 2009;106(52):22540-22545. doi:10.1073/pnas.0911967106.
@Havens_Synthetic_2012
2.	Kyle A. Havens, et al. A Synthetic Approach Reveals Extensive Tunability of Auxin Signaling. Plant Physiology Sep 2012, 160 (1) 135-142; doi: 10.1104/pp.112.202184
@Guseman_Auxin-induced_2015
3.	Guseman JM, et al. Auxin-induced degradation dynamics set the pace for lateral root development. Development (Cambridge, England). 2015;142(5):905-909. doi:10.1242/dev.117234.
@Covington_Circadian_2007
8.	Covington MF, Harmer SL. The Circadian Clock Regulates Auxin Signaling and Responses in Arabidopsis . Weigel D, ed. PLoS Biology. 2007;5(8):e222. doi:10.1371/journal.pbio.0050222.
@Vidal_Nitrate_2010
9.	Vidal EA, Araus V, Lu C, et al. Nitrate-responsive miR393/AFB3 regulatory module controls root system architecture in Arabidopsis thaliana. Proceedings of the National Academy of Sciences of the United States of America. 2010;107(9):4477-4482. doi:10.1073/pnas.0909571107.
@Prigge_Arabidopsis_2016
10.	Prigge MJ, Greenham K, Zhang Y, et al. The Arabidopsis Auxin Receptor F-Box Proteins AFB4 and AFB5 Are Required for Response to the Synthetic Auxin Picloram. G3: Genes|Genomes|Genetics. 2016;6(5):1383-1390. doi:10.1534/g3.115.025585.
@Yu_Untethering_2015
11.	Yu H, Zhang Y, Moss BL, et al. Untethering the TIR1 auxin receptor from the SCF complex increases its stability and inhibits auxin response. Nature plants. 2015;1(3):14030. doi:10.1038/nplants.2014.30.
## Aux/IAA genes

@@

## TPL/TPR genes  
Alex
